A Horrocks-type theorem for orthogonal $K_2$

### 2019-08-20 dl-paper.tex
Added 1 more theorme

The solution of the famous Serre problem on projective modules by D.~Quillen and A.~Suslin was one of the achievements of the mathematics of the 70's,
 which played a significant role in the development of modern algebraic K-theory. In its simplest form it asks whether any projective module over a polynomial ring $k[x_1,\ldots, x_n]$ over a field $k$ is free.

We refer the reader to the textbook~\cite{Lam10} for a more complete introduction to the problem and its history.

@@

There are three main ingredients in the proof of any analogue of Serre's problem: Quillen--Suslin local-global principle, Horrocks theorem, Zariski glueing.
In~\cite{LS17} the authors have already shown that Steinberg groups $\St(\Phi, R)$ satisfy the local-global principle. In this paper we make another step towards
 transferring Tulenbayev's result to Steiberg groups and prove the following theorem.
\begin{theorem*} Suppose $A$ is a local ring with maximal ideal $M$. Then for any $\ell \geq 7$ and $2 \in A^*$ the following commutative square is a pullback in which all arrows are injective:
 \begin{equation} \nonumber
  \xymatrix{ \St(\rD_\ell, A) \ar[r] \ar[d] & \St(\rD_\ell, A[X]) \ar[d] \\ \St(\rD_\ell, A[X\inv]) \ar[r] & \St(\rD_\ell, A[X, X\inv]).}
 \end{equation}
\end{theorem*}
This result should be compared with~\cite[Theorem~1.1]{St-poly}...

The proof of our main result breaks into two parts.

\subsection{Acknowledgements} Theorem 1 is due to the first-named author who acknowledges the financial support of RSCF grant No ???.
Theorem 2 is due to the second-named author who acknowledges the financial support of GazpromNeft and RFBR grant No ???

@@

\section{Proof of the first part of the injectivity theorem} \label{firstPart}
We start this section by recalling basic notation and facts pertaining to the theory of Grothendieck--Witt groups.
%First of all, recall that the stable orthogonal K-groups $\KO_i(R) = \KO_i(\infty, R)$ are precisely the hermitian K-groups ${}_{-1}\!L_i(R)$ introduced by Karoubi in~\cite{Ka80}. We will need yet another redesignation for these groups $\KO_i(R)$, which comes from the theory of {\it higher Grothendieck--Witt groups of schemes}.
Recall that this theory, developed by M.~Schlichting, is a modern broad generalization of the classical hermitian K-theory of rings (see e.\,g.~\cite{Ka80}). We refer the reader to~\cite[\S~2]{FRS12} and~\cite[\S~2]{AF17} for an introduction to Grothendieck--Witt groups.

@@

We are ready to prove our main result.
\begin{theorem} \label{thm:P1glueing}
 Let $\Phi$ be a root system of type $\rA_{\geq 4}, \rD_{\geq 5}$ or $\rE_{6,7,8}$.
 Suppose that the map $j_R \colon \St(\Phi, B) \to \St(\Phi, A[X, X\inv])$ is injective.  
 Then the map $j_-$ is injective and the following commutative square is pullback
 \begin{equation} \nonumber
  \xymatrix{ \St(\Phi, A) \ar[r] \ar[d] & \St(\Phi, A[X]) \ar[d]^{j_+} \\ \St(\Phi, A[X\inv]) \ar[r]^{j_-} & \St(\Phi, A[X, X\inv]).}
 \end{equation}
\end{theorem}
\begin{proof} Since $j_- = j_R j_B^-$, where $j_B^-$ is as in~\cref{prop43}, it suffices to show that $j_B^-$ is injective.
 Set $v_0 = [1, 1, 1] \in \overline{V}_T$ and suppose $g \in \Ker(j_B)$.
 By~\cref{prop43} holds $v_0 = j_B^-(g) \cdot v_0 = [1, g, 1]$ therefore $g \in G_M^0$ and consequently $g = 1$.

@@

>>>>

The solution of the famous Serre problem on projective modules by D.~Quillen and A.~Suslin was one of the achievements of the mathematics of the 70's,
 which played a significant role in the development of modern algebraic K-theory. In its simplest form it asks whether any projective module over a polynomial ring $k[x_1,\ldots, x_n]$ over a field $k$ is free. We refer the reader to the textbook~\cite{Lam10} for a more complete introduction to the problem and its history.

@@

There are three main ingredients in the proof of the $\K_2$-analogue of Serre's problem. These ingredients are Quillen--Suslin local-global principle, Horrocks theorem, Zariski glueing.
In~\cite{LS17} the authors have already shown that Steinberg groups $\St(\Phi, R)$ satisfy the local-global principle. In this paper we make another step towards
 transferring Tulenbayev's result to orthogonal Steiberg groups and prove the following theorem (cf.~\cite[Theorem~5.1]{Tu83})
\begin{theorem}\label{thm:main} Suppose $A$ is a commutative ring in which $2$ is invertible. Then for any $\ell \geq 7$ the following commutative square is a pullback in which all arrows are injective:
\begin{equation} \nonumber
  \xymatrix{ \St(\rD_\ell, A) \ar[r] \ar[d] & \St(\rD_\ell, A[X]) \ar[d] \\ \St(\rD_\ell, A[X\inv]) \ar[r] & \St(\rD_\ell, A[X, X\inv]).}
\end{equation} \end{theorem}
It is easy to see that in the statement of the above theorem the groups $\St(\rD_\ell, -)$  can be replaced with the groups $\K_2(\rD_\ell, -)$. Thus, the above result can be thought of as a direct $\K_2$-analogue of~\cite[Theorem~6.8]{Su82}.

The proof of~\cref{thm:main} goes as follows. After invoking the local-global principle~\cite[Theorem~2]{LS17} the proof reduces to the special case when $A$ is local. Now if $M$ is the maximal ideal of $A$, the proof of the injectivity of $\St(\rD_\ell, A[X]) \to \St(\rD_\ell, A[X, X\inv])$ comes down to proving injectivity of the following two maps
\[ \xymatrix{ \St(\rD_\ell, A[X]) \ar[r] & \St(\rD_\ell, A[X\inv] + M[X]) \ar[r] & \St(\rD_\ell, A[X, X\inv].) }\]
The injectivity of the second map is obtained in~\cref{thm41}. This is the only place in our proof which uses the invertibility of $2$ assumption. This result involves some previously known computations with Grothendieck--Witt groups and also the stabilization theorem for the orthogonal K-theory. The injectivity of the first map is obtained in~\cref{thm:P1glueing}, which is a direct generalization of~\cite[Proposition~4.3]{Tu83}. In the proof of this result we also make explicit many of the details which are left implicit in Tulenbaev's proof.

\subsection{Acknowledgements} Theorem 2 is due to the first-named author who acknowledges the financial support of RSCF grant No ???.
Theorem 3 is due to the second-named author who acknowledges the financial support of JSC Gazpromneft and RFBR grant No 18-31-20044

@@

\section{An injectivity theorem for Steinberg groups} \label{firstPart}
We start this section by recalling basic notation and facts pertaining to the theory of Grothendieck--Witt groups.
%First of all, recall that the stable orthogonal K-groups $\KO_i(R) = \KO_i(\infty, R)$ are precisely the hermitian K-groups ${}_{-1}\!L_i(R)$ introduced by Karoubi in~\cite{Ka80}. We will need yet another redesignation for these groups $\KO_i(R)$, which comes from the theory of {\it higher Grothendieck--Witt groups of schemes}.
Recall that this theory, developed by M.~Schlichting, is a modern broad generalization of the classical hermitian K-theory of rings (see e.\,g.~\cite{Ka80}). We refer the reader to~\cite[\S~2]{FRS12} and~\cite[\S~2]{AF17} for an introduction to Grothendieck--Witt groups.

@@

We are ready for the first main result of this section.
\begin{theorem} \label{thm:P1glueing}
 Let $\Phi$ be a root system of type $\rA_{\geq 4}, \rD_{\geq 5}$ or $\rE_{6,7,8}$.
 Suppose that the map $j_R \colon \St(\Phi, B) \to \St(\Phi, A[X, X\inv])$ is injective.  
 Then the map $j_-$ is injective and the following commutative square is pullback
 \begin{equation} \nonumber
  \xymatrix{ \St(\Phi, A) \ar[r] \ar[d] & \St(\Phi, A[X]) \ar[d]^{j_+} \\ \St(\Phi, A[X\inv]) \ar[r]^{j_-} & \St(\Phi, A[X, X\inv]).}
 \end{equation}
\end{theorem}
\begin{proof} Since $j_- = j_R j_B^-$, where $j_B^-$ is as in~\cref{prop43}, it suffices to show that $j_B^-$ is injective.
 Set $v_0 = [1, 1, 1] \in \overline{V}_T$ and suppose $g \in \Ker(j_B)$.
 By~\cref{prop43} holds $v_0 = j_B^-(g) \cdot v_0 = [1, g, 1]$ therefore $g \in G_M^0$ and consequently $g = 1$.

@@

\begin{proof}[Proof of~\cref{thm:main}]
 The assertion of the theorem follows from~\cref{thm41} and~\cref{thm:P1glueing} in the special case when $A$ is a local ring.

 Now if $g \in \St(\rD_\ell, A[X])$ is such that its image in $\St(\rD_\ell, A[X, X\inv])$ is trivial, then so is its image in all localizations $\St(\rD_\ell, A_M[X, X\inv])$, where $M$ ranges over the maximal ideals of $A$. By the just established local case the images of $g$ in all $\St(\rD_\ell, A_M[X])$ are also trivial hence by the local-global principle \cite[Theorem~2]{LS17} the element $g$ is also trivial.

 The proof of the fact the square of Steinberg groups is pullback is very similar   to the above argument (cf. also with~\cite[Theorem~5.1a]{Tu83}).
\end{proof}
